---
title: Wald tests
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

### Recap

---

### General Wald test

---

### Class activity, Part I

.large[
[https://sta711-s23.github.io/class_activities/ca_lecture_18.html](https://sta711-s23.github.io/class_activities/ca_lecture_18.html)
]

---

### Class activity

```{r, include=F}
library(tidyverse)
titanic <- read.csv("https://sta711-s23.github.io/homework/Titanic.csv")
m1 <- glm(Survived ~ Sex + Age + as.factor(Pclass),
          data = titanic, family = binomial)
```

```{r}
betahat <- m1$coefficients[4:5]
V <- vcov(m1)[4:5, 4:5]
test_stat <- t(betahat) %*% solve(V) %*% betahat
test_stat
```

---

### Class activity

```{r}
betahat <- m1$coefficients[4:5]
V <- vcov(m1)[4:5, 4:5]
test_stat <- t(betahat) %*% solve(V) %*% betahat
test_stat

# rejection region for alpha = 0.05
qchisq(0.05, df=2, lower.tail=F)

# p-value
pchisq(test_stat, df=2, lower.tail=F)
```


---

### A different question

.large[
We have the model
]

$$Y_i \sim Bernoulli(p_i)$$

$$\log \left( \dfrac{p_i}{1 - p_i} \right) = \beta_0 + \beta_1 Sex_i + \beta_2 Age_i + \beta_3 SecondClass_i + \beta_4 ThirdClass_i$$

.large[
We want to test whether there is a difference in the chance of survival for second and third class passengers, holding age and sex fixed. 

.question[
What hypotheses should we test?
]
]

---

### Contrasts

---

### Class activity, Part II

.large[
[https://sta711-s23.github.io/class_activities/ca_lecture_18.html](https://sta711-s23.github.io/class_activities/ca_lecture_18.html)
]

---

### Class activity

```{r}
a <- c(0, 0, 0, -1, 1)
test_stat <- (t(a) %*% coef(m1))/sqrt(t(a) %*% vcov(m1) %*% a)
test_stat
```

---

### Class activity

```{r}
a <- c(0, 0, 0, -1, 1)
test_stat <- (t(a) %*% coef(m1))/sqrt(t(a) %*% vcov(m1) %*% a)
test_stat

# rejection region for alpha = 0.05
qnorm(0.025, lower.tail=F)

# p-value
2*pnorm(abs(test_stat), lower.tail=F)
```


---

### A two-sample test for a difference in means