---
title: Likelihood ratio tests
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

```{r include=F}
library(knitr)
library(tidyverse)
library(MASS)
library(foreign)
library(statmod)
library(gridExtra)

hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

earthquake <- read.csv("https://sta711-s23.github.io/homework/earthquake_small.csv")

```

### Asymptotics of the LRT

---

### Generalization to higher dimensions

---

### Earthquake data

.large[
Data from the 2015 Gorkha earthquake on 211774 buildings, with variables including:

* `Damage`: whether the building sustained any damage (1) or not (0)
* Age: the age of the building (in years)
* `Surface`: a categorical variable recording the surface condition of the land around the building. There are three different levels: n, o, and t
]

---

### Likelihood ratio tests

```{r, output.lines = c(10:16, 21:23)}
m1 <- glm(Damage ~ Age*Surface, data = earthquake,
          family = binomial)
summary(m1)
```

.large[
.question[
We want to test whether the relationship between Age and Damage is the same for all three surface conditions. What hypotheses do we test?
]
]

---

### Likelihood ratio tests

.large[
**Full model:**

```{r, eval=F}
m1 <- glm(Damage ~ Age*Surface, data = earthquake,
          family = binomial)
```

**Reduced model:**

```{r, eval=F}
m2 <- glm(Damage ~ Age + Surface, data = earthquake,
          family = binomial)
```
]

---

### Likelihood ratio tests

```{r, output.lines = c(10:16, 21:23)}
m1 <- glm(Damage ~ Age*Surface, data = earthquake,
          family = binomial)
summary(m1)
```

.large[
.question[
What information replaces $R^2$ and $R^2_{adj}$ in the GLM output?
]
]

---

### Deviance

.large[
**Definition:** The *deviance* of a fitted model with parameter estimates $\widehat{\beta}$ is given by
$$2 \ell(\text{saturated model}) - 2 \ell(\widehat{\beta})$$
]

---

### Residual and null deviance

```{r, output.lines = c(22:23)}
m1 <- glm(Damage ~ Age*Surface, data = earthquake,
          family = binomial)
summary(m1)
```

---

### Comparing deviances

```{r, output.lines = c(22:23)}
m1 <- glm(Damage ~ Age*Surface, data = earthquake,
          family = binomial)
summary(m1)
```

```{r, output.lines = c(20:21)}
m2 <- glm(Damage ~ Age + Surface, data = earthquake,
          family = binomial)
summary(m2)
```

.large[
.question[
How should I use this output to calculate a test statistic?
]
]

---

### Comparing deviances

.large[
```{r}
m1 <- glm(Damage ~ Age*Surface, data = earthquake,
          family = binomial)

m2 <- glm(Damage ~ Age + Surface, data = earthquake,
          family = binomial)

pchisq(m2$deviance - m1$deviance,
       m2$df.residual - m1$df.residual,
       lower.tail = F)
```
]

---

### Summary: LRT for logistic regression
