---
title: Hypothesis testing framework
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

### Recap: Constructing a test

.large[
$$H_0: \theta \in \Theta_0 \hspace{1cm} H_A: \theta \in \Theta_1$$

Observe data $X_1,...,X_n$.

* Calculate a test statistic $T_n = T(X_1,...,X_n)$
* Choose a rejection region $R = \{ (x_1,...,x_n): \text{ reject } H_0\}$
* Reject $H_0$ if $(X_1,...,X_n) \in R$
]

---

### Power function

.large[
Suppose we reject $H_0$ when $(X_1,...,X_n) \in R$. The **power function** $\beta(\theta)$ is
$$\beta(\theta) = P_{\theta}((X_1,...,X_n) \in R)$$
]

---

### Example

.large[
$X_1,...,X_n$ iid from a population with mean $\mu$ and variance $\sigma^2$. 

$$H_0: \mu = \mu_0 \hspace{1cm} H_A: \mu > \mu_0$$
]

---

### Class activity

.large[
$$\beta(\mu) \approx 1 - \Phi\left(z_\alpha - \dfrac{(\mu - \mu_0)}{\sigma/\sqrt{n}} \right)$$

.question[
* Suppose that $\mu_0 = 0$, $n = 100$, and $\sigma = 1$. Make a plot of $\beta(\mu)$ vs. $\mu$ for $\alpha = 0.05$.
* Now consider testing $H_0: \mu \leq \mu_0$ vs. $H_A: \mu > \mu_0$. Will this change our rejection region if we want a size $\alpha$ test?
]
]

---

### Class activity

```{r, echo=F, message=F, warning=F, fig.align='center', fig.width=10, fig.height=4}
library(tidyverse)
library(gridExtra)

mu0 <- 0
mu <- seq(-0.5, 0.5, 0.01)
sigma <- 1
n <- 100
alpha <- 0.05
z <- qnorm(alpha, lower.tail=F)
power <- 1 - pnorm(z - (mu - mu0)/(sigma/sqrt(n)))
p1 <- data.frame(mu, power) %>%
  ggplot(aes(x = mu, y = power)) +
  geom_line(lwd = 1) +
  theme_bw() +
  theme(text = element_text(size = 20))
  

mu <- seq(-0.1, 0.1, 0.01)
power <- 1 - pnorm(z - (mu - mu0)/(sigma/sqrt(n)))
p2 <- data.frame(mu, power) %>%
  ggplot(aes(x = mu, y = power)) +
  geom_line(lwd = 1) +
  theme_bw() +
  theme(text = element_text(size = 20)) +
  geom_hline(yintercept = 0.05, color="blue", lty=2, lwd=1)

grid.arrange(p1, p2, ncol=2)
```

---

### Rejecting $H_0$

.large[
$$H_0: \theta \in \Theta_0 \hspace{1cm} H_A: \theta \in \Theta_1$$

.question[
A hypothesis test rejects $H_0$ if $(X_1,...,X_n)$ is in the rejection region $R$. Are there any issues if we only use a rejection region to test hypotheses?
]
]

---

### p-values

.large[
$$H_0: \theta \in \Theta_0 \hspace{1cm} H_A: \theta \in \Theta_1$$

Given $\alpha$, we construct a rejection region $R$ and reject $H_0$ when $(X_1,...,X_n) \in R$. Let $(x_1,...,x_n)$ be an observed set of data. 

**Definition:** The **p-value** for the observed data $(x_1,...,x_n)$ is the smallest $\alpha$ for which we reject $H_0$.
]

---

### Example

.large[
$X_1,...,X_n$ iid from a population with mean $\mu$ and variance $\sigma^2$. 

$$H_0: \mu = \mu_0 \hspace{1cm} H_A: \mu > \mu_0$$
]

---

### Next steps

.large[
.question[
So far, we have discussed the Wald test in detail. What other hypothesis tests have you seen in statistics courses?
]
]

---

### Class activity

.large[
[https://sta711-s23.github.io/class_activities/ca_lecture_20.html](https://sta711-s23.github.io/class_activities/ca_lecture_20.html)
]

---

### Class activity

```{r, echo=F, message=F, fig.align='center', fig.width=10, fig.height=4}
set.seed(3)
ns <- c(5, 10, 25, 50, 100)
mu0 <- 0
sigma <- 1
alpha <- 0.05
nreps <- 5000
error_known <- rep(0, length(ns))
error_unknown <- rep(0, length(ns))
for(j in 1:length(ns)){
  n <- ns[j]
  test_stats <- rep(0, nreps)
  test_stats_est <- rep(0, nreps)
  for(i in 1:nreps){
    x <- rnorm(n, mu0, sigma)
    test_stats[i] <- (mean(x) - mu0)/(sigma/sqrt(n))
    test_stats_est[i] <- (mean(x) - mu0)/(sd(x)/sqrt(n))
  }
  error_known[j] <- mean(test_stats > qnorm(0.05, lower.tail=F))
  error_unknown[j] <- mean(test_stats_est > qnorm(0.05, lower.tail=F))
}

p1 <- data.frame(n = ns, error = error_known) %>%
  ggplot(aes(x = n, y = error)) +
  geom_point(size = 2) +
  geom_hline(yintercept = 0.05) +
  theme_bw() +
  labs(y = "Type I error") +
  theme(text = element_text(size = 20)) +
  scale_y_continuous(limits = c(0, 0.1))

p2 <- data.frame(n = ns, error = error_unknown) %>%
  ggplot(aes(x = n, y = error)) +
  geom_point(size = 2) +
  geom_hline(yintercept = 0.05) +
  theme_bw() +
  labs(y = "Type I error") +
  theme(text = element_text(size = 20)) +
  scale_y_continuous(limits = c(0, 0.1))

grid.arrange(p1, p2, ncol=2)
```

.large[
.question[
If we reject $H_0: \mu = \mu_0$ when $\frac{\sqrt{n}(\overline{X}_n - \mu_0)}{s} > z_{\alpha}$, why does type I error increase as $n$ decreases?
]
]

