---
title: Convergence of the MLE, Wald tests
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

### Recap: convergence of the MLE

.large[
Under regularity conditions,

* $\widehat{\theta}_n \overset{p}{\to} \theta$ 
* $\sqrt{n}(\widehat{\theta}_n - \theta) \overset{d}{\to} N(0, \mathcal{I}^{-1}_1(\theta))$

]

---

### Regularity conditions

---

### Application to logistic regression

---

### Wald tests for single parameters

.large[
Logistic regression model for the dengue data:

$$Y_i \sim Bernoulli(p_i)$$
]

.large[
$$\log \left( \dfrac{p_i}{1 - p_i} \right) = \beta_0 + \beta_1 WBC_i + \beta_2 PLT_i$$

$H_0: \beta_1 = 0 \hspace{1cm} H_A: \beta_1 \neq 0$

]

---

### Class activity

.large[
[https://sta711-s23.github.io/class_activities/ca_lecture_16.html](https://sta711-s23.github.io/class_activities/ca_lecture_11.html)
]

---

### Class activity

```{r, include=F}
library(tidyverse)
dengue <- read.csv("https://sta712-f22.github.io/homework/dengue.csv")
```

```{r}
m1 <- glm(Dengue ~ WBC + PLT, data = dengue, family = binomial)
X <- model.matrix(m1)
solve(t(X) %*% diag(m1$weights) %*% X)
```

```{r}
vcov(m1)
```

```{r}
summary(m1)$coefficients[,2]^2
```


---

### Class activity

