---
title: Hypothesis testing framework
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

### Summary of Wald tests

.large[
Let $\theta \in \mathbb{R}^p$ be some parameter of interest. We wish to test the hypotheses
$$H_0: C\theta = \gamma_0 \hspace{1cm} H_A: C\theta \neq \gamma_0$$
for some $C \in \mathbb{R}^{q \times p}$. Given an estimator $\widehat{\theta}_n$ such that
$$V_n^{-\frac{1}{2}}(\widehat{\theta}_n - \theta) \overset{d}{\to} N(0, I),$$

the **Wald test** rejects when

$$(C \widehat{\theta}_n - \gamma_0)^T V_n^{-1} (C \widehat{\theta}_n - \gamma_0) > \chi^2_{q, \alpha}$$
]

---

### General framework for hypothesis tests

.large[
**Definition:** Let $\theta \in \mathbb{R}^p$ be some parameter of interest. A **hypothesis** is a statement about $\theta$. The goal of a hypothesis test is to compare two competing hypotheses:

$$H_0: \theta \in \Theta_0 \hspace{1cm} H_A: \theta \in \Theta_1$$
]

---

### Outcomes

.large[
$$H_0: \theta \in \Theta_0 \hspace{1cm} H_A: \theta \in \Theta_1$$

The outcome of the test is a decision to either **reject** $H_0$ or **fail to reject** $H_0$.
]

---

### Constructing a test

.large[
$$H_0: \theta \in \Theta_0 \hspace{1cm} H_A: \theta \in \Theta_1$$
]

---

### Power function

.large[
Suppose we reject $H_0$ when $(X_1,...,X_n) \in R$. The **power function** $\beta(\theta)$ is
$$\beta(\theta) = P_{\theta}((X_1,...,X_n) \in R)$$
]

---

### Example

.large[
$X_1,...,X_n$ iid from a population with mean $\mu$ and variance $\sigma^2$. 

$$H_0: \mu = \mu_0 \hspace{1cm} H_A: \mu > \mu_0$$
]

---

### Class activity

.large[
$$\beta(\mu) \approx 1 - \Phi\left(z_\alpha - \dfrac{(\mu - \mu_0)}{\sigma/\sqrt{n}} \right)$$

.question[
* Suppose that $\mu_0 = 0$, $n = 100$, and $\sigma = 1$. Make a plot of $\beta(\mu)$ vs. $\mu$ for $\alpha = 0.05$.
* Now consider testing $H_0: \mu \leq \mu_0$ vs. $H_A: \mu > \mu_0$. Will this change our rejection region if we want a size $\alpha$ test?
]
]

---

### Class activity

```{r, echo=F, message=F, warning=F, fig.align='center', fig.width=10, fig.height=4}
library(tidyverse)
library(gridExtra)

mu0 <- 0
mu <- seq(-0.5, 0.5, 0.01)
sigma <- 1
n <- 100
alpha <- 0.05
z <- qnorm(alpha, lower.tail=F)
power <- 1 - pnorm(z - (mu - mu0)/(sigma/sqrt(n)))
p1 <- data.frame(mu, power) %>%
  ggplot(aes(x = mu, y = power)) +
  geom_line(lwd = 1) +
  theme_bw() +
  theme(text = element_text(size = 20))
  

mu <- seq(-0.1, 0.1, 0.01)
power <- 1 - pnorm(z - (mu - mu0)/(sigma/sqrt(n)))
p2 <- data.frame(mu, power) %>%
  ggplot(aes(x = mu, y = power)) +
  geom_line(lwd = 1) +
  theme_bw() +
  theme(text = element_text(size = 20)) +
  geom_hline(yintercept = 0.05, color="blue", lty=2, lwd=1)

grid.arrange(p1, p2, ncol=2)
```

---

### Simulation verification

.large[
```{r}
n <- 100
sigma <- 1
mu0 <- 0
nreps <- 5000
test_stats <- rep(0, nreps)
for(i in 1:nreps){
  x <- rnorm(n, mu0, sigma)
  test_stats[i] <- (mean(x) - mu0)/(sigma/sqrt(n))
}
mean(test_stats > qnorm(0.05, lower.tail=F))
```
]

---

### Simulation verification

.large[
```{r}
n <- 100
sigma <- 1
mu0 <- 0
nreps <- 5000
test_stats <- rep(0, nreps)
for(i in 1:nreps){
  x <- rnorm(n, 0.1, sigma)
  test_stats[i] <- (mean(x) - mu0)/(sigma/sqrt(n))
}
mean(test_stats > qnorm(0.05, lower.tail=F))
```
]


