---
title: Inequalities and Asymptotics
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---


### Wald tests for single parameters

.large[
Logistic regression model for the dengue data:

$$Y_i \sim Bernoulli(p_i)$$
]

.large[
$$\log \left( \dfrac{p_i}{1 - p_i} \right) = \beta_0 + \beta_1 WBC_i + \beta_2 PLT_i$$

.question[
Researchers want to know if there is a relationship between white blood cell count and the probability a patient has dengue, after accounting for platelet count. What hypotheses should the researchers test?
]
]

---

### Wald tests for single parameters

```{r include=F}
library(knitr)
hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

dengue <- read.csv("https://sta712-f22.github.io/homework/dengue.csv")
```

.large[
```{r, output.lines = 10:14}
m1 <- glm(Dengue ~ WBC + PLT, data = dengue, 
          family = binomial)
summary(m1)
```
]

---

### What this requires

---

### What we need to do

---

### Markov's inequality

.large[
**Theorem:** Let $Y$ be a non-negative random variable, and suppose that $\mathbb{E}[Y]$ exists. Then for any $t > 0$,
$$P(Y > t) \leq \frac{\mathbb{E}[Y]}{t}$$
]

---

### Chebyshev's inequality

.large[
**Theorem:** Let $Y$ be a random variable, and let $\mu = \mathbb{E}[Y]$ and $\sigma^2 = Var(Y)$. Then
$$P(|Y - \mu| \geq t) \leq \frac{\sigma^2}{t^2}$$

.question[
With your neighbor, apply Markov's inequality to prove Chebyshev's inequality.
]
]

---

### Cauchy-Schwarz inequality

.large[
**Theorem:** For any two random variables $X$ and $Y$,
$$|\mathbb{E}[XY]| \leq \mathbb{E}|XY| \leq (\mathbb{E}[X^2])^{1/2}(\mathbb{E}[Y^2])^{1/2}$$

.question[
**Example:** The *correlation* between $X$ and $Y$ is defined by
$$\rho(X, Y) = \dfrac{Cov(X, Y)}{\sqrt{Var(X)} \sqrt{Var(Y)}}$$
Working with a neighbor, use the Cauchy-Schwarz inequality to prove that $-1 \leq \rho(X, Y) \leq 1$.
]
]

---

### Jensen's inequality

.large[
**Theorem:** For any random variable $Y$, if $g$ is a convex function, then
$$\mathbb{E}[g(Y)] \geq g(\mathbb{E}[Y])$$
]